\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-12 sections/chameleon.tex
Updates from Overleaf

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
In the example of Fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another related form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed. Once a card is selected, the information in the conflicting types block changes according to the change of candidate expression. In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.

@@

Programmers interact with the candidate expression cards by clicking on them to commit the change of meaning or hovering on them to peek at the effect of the change; the effect is reverted once the cursor moves away.

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error fig \ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.

\subsubsection{\textbf{Traditional type errors are biased}}
A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well. In fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect. %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect. 
Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}. This framing is standard practice in most typed languages. However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer. In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.

@@

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are well-typed. Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression. In the editor pane, some error locations change highlight colors based on the updated candidate expression.

@@

Programmers select a candidate expression card by clicking on one card. Alternatively, they can peek at the effect of the change by hovering on one card. The hover effect is reverted once the cursor moves away.
